Selection of Polynomial Features by Genetic Algorithms

### 2013-07-25 texts/paper_spfga-draft.tex
A quantidade de porcaria que os editores são capazes de inventar é assombrosa.

\author[ue,labmag]{Francisco Coelho}
\ead{fc@di.uevora.pt}

\author[fcul,labmag]{João Pedro Neto\corref{jpn}}
\ead{jpn@di.fc.ul.pt}
\cortext[jpn]{Corresponding author}

@@

Polynomials, one of the most studied subjects in mathematics, generalize li\-ne\-ar functions and define, perhaps, the simplest and most used nonlinear models. Applications include colorimetric calibration \citep{mendes2005adaptive}, explicit formulae for turbulent pipe flows \citep{davidson1999method}, computational linguistics \citep{sanchez2009obtaining} and more recently, analytical techniques for cultural heritage materials \citep{csefalvayova2010use}, liquid epoxy molding process \citep{chan2011modeling}, B-spline surface reconstruction \citep{galvez2012iterative}, product design \citep{chan2012development} or forecasting cyanotoxins presence in water reservoirs \citep{garcia2013hybrid}. This not only illustrates the wide spectrum of applications but, additionally, work in each one of these polynomial models uses, at some point, a genetic algorithm.

@@

The proposed method has competitive results comparing with some well-known regression methods. Only Random Forest outperforms \ac{GAPoly} systematically (it also outperforms all the other regression algorithms). One exception --- the artificial dataset that uses a straightforward polynomial relation --- is the Auto MPG dataset where \ac{GAPoly} has comparable results. This is evidence that applying standard genetic operators for polynomial model searching is a viable tool for regression purposes.

>>>>

\author[ue,labmag]{Francisco Coelho\corref{fc}}
\ead{fc@di.uevora.pt}
\cortext[fc]{Corresponding author}

\author[fcul,labmag]{João Pedro Neto}
\ead{jpn@di.fc.ul.pt}

@@

Polynomials, one of the most studied subjects in mathematics, generalize li\-ne\-ar functions and define, perhaps, the simplest and most used nonlinear models. Applications include colorimetric calibration \citep{mendes2005adaptive}, explicit formul\ae\ for turbulent pipe flows \citep{davidson1999method}, computational linguistics \citep{sanchez2009obtaining} and more recently, analytical techniques for cultural heritage materials \citep{csefalvayova2010use}, liquid epoxy molding process \citep{chan2011modeling}, B-spline surface reconstruction \citep{galvez2012iterative}, product design \citep{chan2012development} or forecasting cyanotoxins presence in water reservoirs \citep{garcia2013hybrid}. This not only illustrates the wide spectrum of applications but, additionally, work in each one of these polynomial models uses, at some point, a genetic algorithm.

@@

The proposed method has competitive results comparing with some well-known regression methods. Only Random Forest outperforms \ac{GAPoly} systematically (it also outperforms all the other regression algorithms). One exception --- besides the artificial dataset that uses a straightforward polynomial relation --- is the Auto MPG dataset where \ac{GAPoly} has comparable results. This is evidence that applying standard genetic operators for polynomial model searching is a viable tool for regression purposes.

### 2013-07-25 texts/submission_01-prl/paper_spfga-final.tex
A quantidade de porcaria que os editores são capazes de inventar é assombrosa.

\documentclass[draft,authoryear,3p,times,twocolumn]{elsarticle}

@@

%
%	ACRONYMS
%
\usepackage{acronym}
\acrodef{GA}{Genetic algorithms}
\acrodef{rmse}[error]{root-mean-square error}
\acrodef{GAPoly}[{\sc GAPoly}]{Genetic Algorithms for Polynomials}
%

@@

%
%	TIKZ AND RELATED STUFF
%
\usepackage{pgfplots}
\pgfplotsset{compat=newest}
\usepgfplotslibrary{statistics}
%
%	avoid processing figures every time
%\usepgfplotslibrary{external}
%\tikzexternalize
% ENABLE NEXT LINE TO TYPESET FIGURES
%\tikzset{external/force remake}
%

@@

\item Find the fittest set of monomials, \emph{i.e.} the polynomial that minimizes the \ac{rmse} on the same dataset;
\end{enumerate}
%
More precisely, concerning the first problem, let $\mathcal{D}$ be a dataset with $n$ observations of variables $Y, X_1,\ldots,X_m$ and $\mathcal{P} = \left\lbrace q_1,\ldots, q_k\right\rbrace$ a set of $k$ monomial expressions over $X_1,\ldots,X_m$. Define the hypothesis\footnote{The expression $q|_{X=x}$ reads ``\emph{replace all instances of $X$ by $x$ in $q$}''.}
%
\begin{eqnarray*}
h_{\Theta,\mathcal{P}}\left(x_1,\ldots,x_m\right) &=& \sum_{j = 1}^k \theta_j q_j|_{X_i=x_i,\forall 1 \leq i \leq m}
\end{eqnarray*}
%
and let the cost
%
\begin{eqnarray}
J_{\textrm{fit}}\left(\Theta;\mathcal{P},\mathcal{D}\right)  & =  \nonumber \\
\sqrt{\frac{1}{n}\sum_{i=1}^n \left( y\sample{i} - h_{\Theta,\mathcal{P}}\left( x_1\sample{i},\ldots,x_m\sample{i} \right) \right)^2 }\label{eq:rmse}
\end{eqnarray}
%
be the usual \acf{rmse} function. Now the first problem can be stated as:
%
\begin{quote}
Given a dataset $\mathcal{D}$ and a set of monomials $\mathcal{P}$, find parameters $\Theta$ that minimize $J_{\textrm{fit}}\left(\Theta;\mathcal{P},\mathcal{D}\right)$.
\end{quote}
%
It turns out that this problem can be solved as a usual linear regression problem by expanding $\mathcal{D}$ with columns that replicate the monomials in $\mathcal{M}$.

@@

\begin{figure*}[tb]\begin{center}
\begin{tabular}{cc}
\begin{tikzpicture}
\begin{axis}[
	width=0.49\textwidth,height=4cm,
	boxplot/draw direction=x,
	ytick={1,2,3,4,5,6,7},
	yticklabels={0.7, 0.8, 0.9, 1.0, 1.1, 1.2, 1.3},
	title = {Dataset: \textsc{Abalone}},
	xlabel = {Error distribution},
	ylabel={Regularization weight ($\lambda$)},
	xmax=2.7,
	]	
	%\pgfplotsset{boxplot/every box/.append style={fill=gray!50!white}}
	\addplot[boxplot] table [y index=0] {data/figure1.dat};
	\addplot[boxplot][fill=gray!50!white] table [y index=2] {data/figure1.dat};
	\addplot[boxplot] table [y index=4] {data/figure1.dat};
	\addplot[boxplot] table [y index=6] {data/figure1.dat};		
	\addplot[boxplot] table [y index=8] {data/figure1.dat};
	\addplot[boxplot] table [y index=10] {data/figure1.dat};
	\addplot[boxplot] table [y index=12] {data/figure1.dat};
\end{axis}	
\end{tikzpicture}
%
% RIGHT PANE
&
%
%
\begin{tikzpicture}
\begin{axis}[
	width=0.49\textwidth,height=4cm,
	boxplot/draw direction=x,
	ytick={1,2,3,4,5,6,7},
	%yticklabels={0.7, 0.8, 0.9, 1.0, 1.1, 1.2, 1.3},
	yticklabels={},
	title = {Dataset: \textsc{Auto MPG}},
	xlabel = {Error distribution},
	ylabel={},
	]	
	%\pgfplotsset{boxplot/every box/.append style={fill=gray!50!white}}
	\addplot[boxplot] table [y index=0] {data/lambda.errors.Auto-Mpg.dat};
	\addplot[boxplot][fill=gray!50!white] table [y index=2] {data/lambda.errors.Auto-Mpg.dat};
	\addplot[boxplot] table [y index=4] {data/lambda.errors.Auto-Mpg.dat};
	\addplot[boxplot] table [y index=6] {data/lambda.errors.Auto-Mpg.dat};		
	\addplot[boxplot] table [y index=8] {data/lambda.errors.Auto-Mpg.dat};
	\addplot[boxplot] table [y index=10] {data/lambda.errors.Auto-Mpg.dat};
	\addplot[boxplot] table [y index=12] {data/lambda.errors.Auto-Mpg.dat};

\end{axis}
\end{tikzpicture}
\end{tabular}

@@

The results were found using \texttt{R} programming language~\citep{R}. To compare this paper's proposed algorithm we applied the exact same train and test samples using several well-known learning algorithms for regression, namely: Linear Regression, Support Vector Machines~\citep{Meyer12}, Regression Trees~\citep{Therneau13}, Conditional Inference Trees~\citep{Hothorn06, Strobl07, Strobl08} and Random Forests~\citep{Liaw02}

@@

\begin{figure}[tb]\begin{center}
\begin{tikzpicture}
\begin{axis}[width=0.48\textwidth,height=4cm,
	boxplot/draw direction=x,
	ytick={1,2,3,4,5,6},
	%yticklabels={GA.Poly,Rnd.For,Lin.Reg,SVM,RPART,CI.Tree},
	ylabel={Method index},
	xmin=-25,
	xmax=325,
	xlabel={Error distribution},
	title={Dataset: \textsc{Artificial}},
	]
	%\pgfplotsset{boxplot/every box/.append style={fill=gray!50!white}}
	\addplot[boxplot][fill=gray!50!white] table [y index=0] {data/Articial_dataset.dat};
	\addplot[boxplot] table [y index=4] {data/Articial_dataset.dat};
	\addplot[boxplot] table [y index=1] {data/Articial_dataset.dat};
	\addplot[boxplot] table [y index=2] {data/Articial_dataset.dat};
	\addplot[boxplot] table [y index=3] {data/Articial_dataset.dat};
	\addplot[boxplot] table [y index=5] {data/Articial_dataset.dat};

\end{axis}
\end{tikzpicture}

@@

\newcommand{\axpl}[3]{
\begin{axis}[width=0.49\textwidth,height=4cm,
	ticks=major,
	boxplot/draw direction=x,
	ytick={1,2,3,4,5,6},
	%yticklabels={GA.Poly, Lin.Reg, SVM, RPART,Rnd.For,CI.Tree},
	title={Dataset: \textsc{#2}},
	ylabel={Method index},
	xlabel={Error distribution},
	separate axis lines,
	axis lines = box,
%	axis y line = #3,
	]
	%\pgfplotsset{boxplot/every box/.append style={fill=gray!50!white}}
	\addplot[boxplot][fill=gray!50!white] table [y index=0] {data/#1.dat};
	\addplot[boxplot] table [y index=4] {data/#1.dat};
	\addplot[boxplot] table [y index=1] {data/#1.dat};
	\addplot[boxplot] table [y index=2] {data/#1.dat};
	\addplot[boxplot] table [y index=3] {data/#1.dat};
	\addplot[boxplot] table [y index=5] {data/#1.dat};

\end{axis}
}

\begin{figure*}[tb]\begin{center}
\begin{tabular}{cc}
\begin{tikzpicture}
	\axpl{Housing300_lambda0.8_25runs}{Housing}{left}
\end{tikzpicture}
%
&
%
\begin{tikzpicture}
	\axpl{Abalone300_lambda0.8_25runs}{Abalone}{right}
\end{tikzpicture}
%
\\
%
\begin{tikzpicture}
	\axpl{Auto-Mpg300_lambda0.8_25runs}{Auto MPG}{left}
\end{tikzpicture}
%
&
%
\begin{tikzpicture}
	\axpl{Kinematics300_lambda0.8_25runs}{Kinematics}{right}
\end{tikzpicture}

\end{tabular}

@@

\begin{figure*}[tb]
\begin{center}
\begin{tikzpicture}

\begin{axis}[
	width=0.98\textwidth,
	height=4cm,
	xmin=-5,
	xmax=245,
	%line width=1pt,
	xlabel={Number of iterations},
	ylabel={Error},
	title={Error evolution with number of iterations (Dataset: \textsc{Abalone})},
	]

	\addplot[solid,color=black] table [x expr=\coordindex, y index = 0] {data/lambdas.dat};\addlegendentry{$\lambda=0.975$};
	\addplot[dashed,color=black] table [x expr=\coordindex, y index = 1] {data/lambdas.dat};\addlegendentry{$\lambda=1.000$};
\end{axis}
\end{tikzpicture}
\caption{Error progress for Abalone dataset during a single execution of the genetic algorithm. The figure shows the fitness evolution for two different regularization values. The population for both consisted of $200$ polynomials. The error values seem to stabilize around iteration~250.}
\label{Abalone_fitnessProgress}

>>>>

\documentclass[final,authoryear,3p,times,twocolumn]{elsarticle}

@@

%
%	ACRONYMS
%
\usepackage{acronym}
\acrodef{GA}{Genetic algorithms}
\acrodef{rmse}[error]{root-mean-square error}
\acrodef{GAPoly}[\textsc{GAPoly}]{Genetic Algorithms for Polynomials}
%

@@

@@

\item Find the fittest set of monomials, \emph{i.e.} the polynomial that minimizes the \ac{rmse} on the same dataset;
\end{enumerate}
%
More precisely, concerning the first problem, let $\mathcal{D}$ be a dataset with $n$ observations of variables $Y, X_1,\ldots,X_m$ and $\mathcal{P} = \left\lbrace q_1,\ldots, q_k\right\rbrace$ a set of $k$ monomial expressions over $X_1,\ldots,X_m$. Define the hypothesis\footnote{The expression $q|_{X=x}$ reads ``\emph{replace all instances of $X$ by $x$ in $q$}''.}
%
\begin{eqnarray*}
h_{\Theta,\mathcal{P}}\left(x_1,\ldots,x_m\right) &=& \sum_{j = 1}^k \theta_j q_j|_{X_i=x_i,\forall 1 \leq i \leq m}
\end{eqnarray*}
%
and let the cost
%
\begin{eqnarray}
J_{\textrm{fit}}\left(\Theta;\mathcal{P},\mathcal{D}\right)  & =  \nonumber \\
\sqrt{\frac{1}{n}\sum_{i=1}^n \left( y\sample{i} - h_{\Theta,\mathcal{P}}\left( x_1\sample{i},\ldots,x_m\sample{i} \right) \right)^2 }\label{eq:rmse}
\end{eqnarray}
%
be the usual \acf{rmse} function. Now the first problem can be stated as:
%
\emph{Given a dataset $\mathcal{D}$ and a set of monomials $\mathcal{P}$, find parameters $\Theta$ that minimize $J_{\textrm{fit}}\left(\Theta;\mathcal{P},\mathcal{D}\right)$.}

%
It turns out that this problem can be solved as a usual linear regression problem by expanding $\mathcal{D}$ with columns that replicate the monomials in $\mathcal{M}$.

@@

%
\begin{figure*}[tb]\begin{center}
\begin{tabular}{cc}
\includegraphics[width=0.49\textwidth]{figure_1a.pdf}
%
% RIGHT PANE
&
%
%
\includegraphics[width=0.49\textwidth]{figure_1b.pdf}
\end{tabular}

@@

The results were found using \texttt{R} programming language~\citep{R}\footnote{The datasets and the \texttt{R} code used to produce the results and plots in this paper are available online at \url{https://github.com/jpneto/GenAlgPoly}.}. To compare this paper's proposed algorithm we applied the exact same train and test samples using several well-known learning algorithms for regression, namely: Linear Regression, Support Vector Machines~\citep{Meyer12}, Regression Trees~\citep{Therneau13}, Conditional Inference Trees~\citep{Hothorn06, Strobl07, Strobl08} and Random Forests~\citep{Liaw02}

@@

\begin{figure}[tb]\begin{center}
\includegraphics[width=0.48\textwidth]{figure_2.pdf}

@@

\begin{figure*}[tb]\begin{center}
\begin{tabular}{cc}
\includegraphics[width=0.48\textwidth]{figure_3a.pdf}
%
&
%
\includegraphics[width=0.48\textwidth]{figure_3b.pdf}
%
\\
%
\includegraphics[width=0.48\textwidth]{figure_3c.pdf}
%
&
%
\includegraphics[width=0.48\textwidth]{figure_3d.pdf}
%
\end{tabular}

@@

\begin{figure*}[tb]
\begin{center}
\includegraphics[width=0.98\textwidth]{figure_4.pdf}
\caption{Error progress for Abalone dataset during a single execution of the genetic algorithm. The figure shows the fitness evolution for two different regularization values. The population for both consisted of $200$ polynomials. The error values seem to stabilize around iteration~250.}
\label{Abalone_fitnessProgress}
